Approximating the matrix exponential derivative of the CTMC generator matrix, with application to global SARS-CoV-2

### 2023-03-14 manuscript/main.tex
money
more writing

%\newcommand{\TTTheta}{\boldsymbol{\Theta}}
%\newcommand{\tr}{\mbox{tr}}
%\newcommand{\X}{\mathbf{X}}
%\renewcommand{\u}{\mathbf{u}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\QQ}{\mathbf{Q}}
\newcommand{\DD}{\mathbf{D}}
\newcommand{\MM}{\mathbf{M}}
\newcommand{\JJ}{\mathbf{J}}
\newcommand{\II}{\mathbf{I}}
\newcommand{\RR}{\mathbf{R}}
%\newcommand{\haar}{\mathcal{H}}
%\newcommand{\orthog}{\mathcal{O}}
%\newcommand{\SSigma}{\boldsymbol{\Sigma}}
\newcommand{\Zero}{\boldsymbol{0}}
\newcommand{\ttheta}{\boldsymbol{\theta}}
\newcommand{\Llambda}{\boldsymbol{\Lambda}}
\newcommand{\vecc}{\mbox{vec}}
\newcommand{\CC}{\mathbf{C}}
\newcommand{\BB}{\mathbf{B}}
\renewcommand{\AA}{\mathbf{A}}
\newcommand{\GG}{\mathbf{G}}
\newcommand{\mom}{\boldsymbol{\xi}}

@@

@@

A small body of work considers another approach by replacing $\nabla \log \pi (\ttheta)$ with a suitable approximation $\widehat{\nabla} \log \pi (\ttheta)$ and recognizing that the modified \eqref{eq:leap} continues to satisfy path reversibility and volume preservation, two essential ingredients for well-specified HMC. It follows immediately that the resulting `surrogate trajectory HMC' continues to sample the correct target distribution $\pi(\cdot)$.  Nonetheless, the acceptance rates and overall efficiency of such samplers may suffer when approximations are poor.   The majority of surrogate HMC methods first obtain a small sample of exact gradient evaluations and then use some model to interpolate: \citet{zhang2017precomputing} assume the approximate gradient follows a piecewise constant form across a grid; \citet{rasmussen2003gaussian,lan2016emulation} construct approximations using Gaussian processes; and \citet{zhang2017hamiltonian,li2019neural} do the same using neural networks.  But an even simpler approach to surrogate HMC may be appropriate when gradients have series representations.   \textcolor{red}{MAGEE} consider phylogenetic models that involve CTMC priors and show that the first-order approximation \eqref{eq:firstOrder} of the matrix exponential derivative \eqref{eq:deriv} performs remarkably well within surrogate HMC, achieving an over 30-fold efficiency gain compared to random-walk Metropolis for a 14-state model with over 180 model parameters.  Here, we demonstrate similar strong performance of the first-order approximation within surrogate HMC for a 44-state model with almost 1900 model parameters.

>>>>

%\newcommand{\TTTheta}{\boldsymbol{\Theta}}
%\newcommand{\tr}{\mbox{tr}}
%\newcommand{\X}{\mathbf{X}}
%\renewcommand{\u}{\mathbf{u}}
\newcommand{\Y}{\mathbf{Y}}
\newcommand{\QQ}{\mathbf{Q}}
\newcommand{\DD}{\mathbf{D}}
\newcommand{\MM}{\mathbf{M}}
\newcommand{\JJ}{\mathbf{J}}
\newcommand{\II}{\mathbf{I}}
\newcommand{\RR}{\mathbf{R}}
%\newcommand{\haar}{\mathcal{H}}
%\newcommand{\orthog}{\mathcal{O}}
%\newcommand{\SSigma}{\boldsymbol{\Sigma}}
\newcommand{\Zero}{\boldsymbol{0}}
\newcommand{\ttheta}{\boldsymbol{\theta}}
\newcommand{\Llambda}{\boldsymbol{\Lambda}}
\newcommand{\vecc}{\mbox{vec}}
\newcommand{\CC}{\mathbf{C}}
\newcommand{\BB}{\mathbf{B}}
\renewcommand{\AA}{\mathbf{A}}
\newcommand{\GG}{\mathbf{G}}
\newcommand{\mom}{\boldsymbol{\xi}}
\newcommand{\ppi}{\boldsymbol{\pi}}
\renewcommand{\P}{\mathbf{P}}

@@

\subsection{Surrogate trajectory HMC}

@@

A small body of work considers another approach by replacing $\nabla \log \pi (\ttheta)$ with a suitable approximation $\widehat{\nabla} \log \pi (\ttheta)$ and recognizing that the modified \eqref{eq:leap} continues to satisfy path reversibility and volume preservation, two essential ingredients for well-specified HMC. It follows immediately that the resulting `surrogate trajectory HMC' continues to sample the correct target distribution $\pi(\cdot)$.  Nonetheless, the acceptance rates and overall efficiency of such samplers may suffer when approximations are poor.   The majority of surrogate HMC methods first obtain a small sample of exact gradient evaluations and then use some model to interpolate: \citet{zhang2017precomputing} assume the approximate gradient follows a piecewise constant form across a grid; \citet{rasmussen2003gaussian,lan2016emulation} construct approximations using Gaussian processes; and \citet{zhang2017hamiltonian,li2019neural} do the same using neural networks.  But an even simpler approach to surrogate HMC may be appropriate when gradients have series representations.   \textcolor{red}{MAGEE} consider phylogenetic models that involve CTMC priors and show that the first-order approximation \eqref{eq:firstOrder} of the matrix exponential derivative \eqref{eq:deriv} performs remarkably well within surrogate HMC, achieving an over 30-fold efficiency gain compared to random-walk Metropolis for a 14-state model with over 180 model parameters.  Here, we demonstrate similar strong performance of the first-order approximation within surrogate HMC for a 44-state model with almost 1,900 model parameters.

\subsection{Phylogeographic CTMC model}

Start with a (possibly unknown) rooted and bifurcating phylogenetic tree $\phylogeny$ consisting of $N$ leaf nodes that correspond to observed biological specimens and $N-1$ internal nodes that correspond to unobserved ancestors.  The tree also contains $2N-2$ branches of length $t_v$ connecting each child node $v$ to its parent $u$.  
 Given $\phylogeny$, we model the evolution of $S$ characters along each branch of the tree according to a CTMC model with $S\times S$ generator matrix $\QQ$ and stationary distribution $\ppi=(\pi_1,\dots,\pi_S)=\lim_{t\rightarrow \infty}e^{t\QQ}\widetilde{\ppi}$, for $\widetilde{\ppi}$ for any arbitrary probability vector.   Examples of characters are the $S=4$ nucleotides within a set of aligned genome sequences  \citep{jukes1969evolution} and the set of $S=15$ geographic regions visited by an influenza subtype \citep{lemey2014unifying}.   Depending on the given problem, we may scale $t_v$ to be raw time (e.g., years in Section \ref{sec:sars}) or the expected number of substitutions with respect to $\ppi$.   In the former case, one may augment the model with a rate scalar $\gamma$ that modulates the expected number of substitutions across all branches, and the finite-time transition probability matrix along branch $v$ becomes $\P_v:=e^{\gamma t_v \QQ}$.

\subsection{Bayesian analysis of SARS-CoV-2 contagion}\label{sec:sars}

\section{Discussion}

\section*{Acknowledgments}

AJH is supported by a gift from the Karen Toffler Charitable Trust and by grants NIH K25 AI153816, NSF DMS 2152774 and NSF DMS 2236854.
